Exploring Local Explanations of Non-linear Models Using Animated Linear Projections

### 2022-01-15 _paper.rmd

<!-- SHAP -->
SHaply Additive exPlanations (SHAP) approximates the feature importance in the vicinity of one instance conceptually by examining the effect of other features on the contribution of the feature of interest on predicting the response. These explanations almost all point to @shapley_value_1953's method to evaluate an individual's contribution to cooperative games by permuting the players that contributing to the score. @strumbelj_efficient_2010 introduced SHAP for local explanations in ML models. While the models themselves are invariant to the order of the predictors, the additive explanations are not. The SHAP values are the mean contributions over some subset of these permutations. This is also related to partial dependence plots [@molnar_interpretable_2020], used to explain the effect of a feature by predicting the response for a range of values on this feature, after fixing the value of all other features to their mean. Partial dependence plots are a global approximation of the feature importance, while SHAP is specific to one instance. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in @wickham_visualizing_2015, which helps to understand the relative importance of each feature in the context of all other candidate features.

<!-- tree SHAP -->
For our application, we use _tree SHAP_, a variant of SHAP that enjoys a lower computational complexity [@lundberg_consistent_2018]. Instead of aggregating over permutations of the feature order, tree SHAP approximates instance-level feature importance by exploring the structure of the decision trees. Tree SHAP is only compatible with tree-based models; we illustrate random forests. The following section will use normalized explanations as to the starting projection basis (call this the _attribution projection_) to further scrutinize the explanation.

```{r shapdistrbd, echo=F, fig.align='center', out.width="60%", fig.cap = "Illustration of the use of SHAP values for a random forest model for salaries of FIFA 2020 players based on nine predictors corresponding to different skills. A star offensive and defensive player, Messi and a defensive player are compared, L. Messi and V van Dijk respectively. Panel (a) is the break-down plot showing one additive feature-attribution. These attributions are sensitive to the order of the features. Panel (b) shows the distribution of attribution for each feature across 25 such permutation of the order of predictors, with the mean displayed as a dot, for each players. Offense and movement are important for Messi but not van Dijk, and conversely, defense and power are important for van Dijk but not Messi."}
knitr::include_graphics("./figures/shap_distr_bd.png")
```

>>>>

<!-- SHAP -->
SHaply Additive exPlanations (SHAP) quantifies the feature contributions of one instance, by examining the effect of other features on the predictions. The explanations of SHAP almost all refer to @shapley_value_1953's method to evaluate an individual's contribution to cooperative games by assessing the performance of this player in the presence or absence of other players. @strumbelj_efficient_2010 introduced the use of SHAP for local explanations in ML models. <!--While the models themselves are invariant to the order of the predictors, the additive explanations are not.--> The SHAP values are the mean contributions over some subset of permutations of features. The approach is  related to partial dependence plots [@molnar_interpretable_2020], used to explain the effect of a feature by predicting the response for a range of values on this feature, after fixing the value of all other features to their mean. Partial dependence plots are a global approximation of the feature importance, while SHAP is specific to one instance. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in @wickham_visualizing_2015, which helps to understand the relative importance of each feature in the context of all other candidate features.

<!-- tree SHAP -->
For our application, we use _tree SHAP_, a variant of SHAP that enjoys a lower computational complexity [@lundberg_consistent_2018]. Instead of aggregating over permutations of the feature order, tree SHAP approximates instance-level feature importance by exploring the structure of the decision trees. Tree SHAP is only compatible with tree-based models; we illustrate random forests. The following section will use normalized SHAP values as a projection basis (call this the _attribution projection_) which will have coefficients varied in order to further scrutinize the feature contributions.

```{r shapdistrbd, echo=F, fig.align='center', out.width="60%", fig.cap = "Illustration of the use of SHAP values for a random forest model for salaries of FIFA 2020 players based on nine predictors corresponding to different skills. A star offensive and defensive player are compared, L. Messi and V. van Dijk, respectively. Panel (a) is the break-down plot showing one additive feature-attribution.  Panel (b) shows the distribution of attribution for each feature across 25 permutations of predictors, with the mean displayed as a dot, for each players. Offense and movement are important for Messi but not van Dijk, and conversely, defense and power are important for van Dijk but not Messi."}
knitr::include_graphics("./figures/shap_distr_bd.png")
```
